---
title: "Century of Death"
output:
  html_document:
    css: style/branding.css
    includes:
      in_header: header_branding.html
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE, message=FALSE, fig.width=11, fig.align = "center")

# To-do list:
## Add 1930 (1930 vital statistics report unavailable).
## Jaspreet and Michael will talk about bringing in the ccb data with code.
## Get clinical consultation on this project.
## Figure out the "kidney disease and diabetes" issue. 
## Review and edit bullets carefully, making sure to avoid syntax/grammer that is not smooth. Remove any comments that show speculation.
## Move "non-core" charts to the end, such that the top bullets, titles, and charts is the "story" we would share with others. -IN PROGRESS-
## Make a set of bullets (or prose; whatever you prefer) that succinctly and clearly describes the process of collecting and extracting the data, mapping the "raw" disease names to the specific level, specific level to display level, and display level to top level--for interest, give a couple of example of "colorful/odd" "raw" conditions names from back in the day.
## explore how what is currently the third chart can be even more "legible" with wider bars (?), different colors, and/or other things you can think of--this is a key chart.
## Add to the bottom of the markdown the 1-3 key references of work most related to this project.
## Sort out Kidney disease in Other communicable

# Completed Items:
## Add numbers to the chart titles. -COMPLETED-
## Kidney disease should be classified as other chronic. -COMPLETED-
## Verify that 2019 and 2020 sorting are identical. -COMPLETED-
## Add 2019 population data. -COMPLETED-
## Add 1980. -COMPLETED-
## Align trend line plots side-to-side. -COMPLETED-
## Add 1919 (low priority). -COMPLETED-
## Address potential miss classification error around 1940-1950. This can be seen in the line plots. -COMPLETED- [ISSUE CAUSED BY 1940]

## Jaspreet - Flextable

# NOTABLE YEARS:

## 1987 - This was the year AZT, the first drug for treating AIDS, gained FDA approval.

```

```{=html}
<style type="text/css">

.main-container {
  max-width: 95% !important;
  margin-left: auto;
  margin-right: auto;
}


body { 
  font-size: 16px;
  color: black;
}

h1.title {
  font-size: 26px;
  color: DarkBlue;
  text-align: center;
  text-decoration: none;
}

h3.subtitle {
  font-size: 16px;
  color: DarkBlue;
  text-align: center;
  text-decoration: none;
}

h4.author {
  font-size: 16px;
  color: DarkBlue;
  text-align: center;
  text-decoration: none;
}

 

h1 { /* Header 1 */
  font-size: 22px;
  color: black;
  text-decoration: underline;
}


h2 {
  font-size: 22px;
  color: DarkBlue;
  margin-left: 0px;
}

h3 {
  font-size: 16px;
  font-face: bold;
  margin-left: 20px;
  color: black;
}


h4 {
  font-size: 16px;
  color: DarkBlue;
}

h5 {
  font-size: 14px;
  color: black;
}



ul {
  display: block;
  font-size: 16px;
  margin-top: 0;
  margin-bottom: 0;
  margin-left: 0;
  margin-right: 0;
  color: black;
}


body{
  font-family: Helvetica !important;
}


</style>
```
```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
library(forcats)
library(knitr)
library(gridExtra)

# Set theme

cdph1 <- 
  theme_bw() + 
  theme(panel.grid.minor = element_blank(), 
        plot.title       = element_text(size = 16, color="darkblue"),
        axis.title       = element_text(size = 14), 
        axis.text  = element_text(size = 14)
        ) 

theme_set(cdph1)

myLineSize <- 1 

y1900 <- read_excel("Century of death DATA.xlsx",sheet="y1900") %>% mutate(year=1900)
y1910 <- read_excel("Century of death DATA.xlsx",sheet="y1910") %>% mutate(year=1910)
y1911 <- read_excel("Century of death DATA.xlsx",sheet="y1911") %>% mutate(year=1911)
y1912 <- read_excel("Century of death DATA.xlsx",sheet="y1912") %>% mutate(year=1912)
y1917 <- read_excel("Century of death DATA.xlsx",sheet="y1917") %>% mutate(year=1917)
y1918 <- read_excel("Century of death DATA.xlsx",sheet="y1918") %>% mutate(year=1918)
y1919 <- read_excel("Century of death DATA.xlsx",sheet="y1919") %>% mutate(year=1919)
y1920 <- read_excel("Century of death DATA.xlsx",sheet="y1920") %>% mutate(year=1920)
y1929 <- read_excel("Century of death DATA.xlsx",sheet="y1929") %>% mutate(year=1929)
y1939 <- read_excel("Century of death DATA.xlsx",sheet="y1939") %>% mutate(year=1939)
y1940 <- read_excel("Century of death DATA.xlsx",sheet="y1940") %>% mutate(year=1940)
y1950 <- read_excel("Century of death DATA.xlsx",sheet="y1950") %>% mutate(year=1950)
y1960 <- read_excel("Century of death DATA.xlsx",sheet="y1960") %>% mutate(year=1960)
y1970 <- read_excel("Century of death DATA.xlsx",sheet="y1970") %>% mutate(year=1970)
y1980 <- read_excel("Century of death DATA.xlsx",sheet="y1980") %>% mutate(year=1980)
y1990 <- read_excel("Century of death DATA.xlsx",sheet="y1990") %>% mutate(year=1990)
y2000 <- read_excel("Century of death DATA.xlsx",sheet="y2000") %>% mutate(year=2000)
y2010 <- read_excel("Century of death DATA.xlsx",sheet="y2010") %>% mutate(year=2010)
y2018 <- read_excel("Century of death DATA.xlsx",sheet="y2018") %>% mutate(year=2018)
y2019 <- read_excel("Century of death DATA.xlsx",sheet="y2019") %>% mutate(year=2019)
y2020 <- read_excel("Century of death DATA.xlsx",sheet="y2020") %>% mutate(year=2020)



workDat0 <- bind_rows(y1900, y1910, y1911, y1912, y1917, y1918, y1919, y1920, y1929, y1939, y1940, y1950, y1960, y1970, y1980, y1990, y2000, y2010, y2018, y2019, y2020)



# Deaths and Population sheet
popDat <- read_excel("Century of death DATA.xlsx",sheet="Deaths and Popu.") %>%
  rename(year = 1, nDeaths = 2, population = 3, cDeathRate = 4) %>%
  select(-5)

#CCB Data Analysis
y2000ccb <- read_excel("Century of death CCB Mapping.xlsx",sheet="ccb_cause_to_100_year_causes",range="$C$1:$G$60") %>%
  rename(Number = deaths_2000) %>% 
  mutate(year=2000.2)
y2010ccb <- read_excel("Century of death CCB Mapping.xlsx",sheet="ccb_cause_to_100_year_causes")[,c(3,4,5,6,8)] %>%
  rename(Number = deaths_2010) %>% 
  mutate(year=2010.2)
y2018ccb <- read_excel("Century of death CCB Mapping.xlsx",sheet="ccb_cause_to_100_year_causes")[,c(3,4,5,6,9)] %>%
  rename(Number = deaths_2018) %>% 
  mutate(year=2018.2)



workDat1 <- bind_rows(y2000, y2000ccb, y2010, y2010ccb, y2018, y2018ccb)
```

# Background & Purpose:

-   Overall death rates and the leading causes of death have changed greatly over the past 100 years. In general, death rates have decreased and leading causes of death have shifted from infectious to chronic.
-   In the big picture these changes are due to improvements in sanitation and hygiene, clean air and water, availability of good food, vaccines, antibiotics, other medicine and advances in health care, and a range of highly effective public health efforts.
-   While the overall improvements and changes in causes of death are well known, collecting, collating and displaying these data carefully and in detail can provide further actionable insights for public health professionals to make informed decisions.
-   Towards this end, a range of paper (scanned into pdf), machine readable query-based data, and our own data systems were reviewed, standardized, synthesized, and ultimately visualized as presented below. 

<br>

# Methods:

-   Data for the years 1900-1990 were extracted from historical vital statistics reports published by the CDC National Center for Health Statistics. These reports were available as pdf scans of printed reports, and the data was thus manually extracted from these documents into a digital Excel spreadsheet.

-   Data for years 2000 to 2018 were initially extracted electronically from the CDC WONDER data query system. Data for these same years were subsequently extracted from the CDPH Fusion Center California Community Burden of Disease and Cost Engine (CCB) system and compared to the WONDER data. Based on detailed review and comparison of these two sources we determined that 1) total numbers of deaths and number of deaths based on broad cause of death groupings were consistent and 2) the CCB groupings and data system provided more detail and flexibility consistent with project goals. Therefore, data for 2000 and subsequent years were extracted, via computer code, from the CCB system.

-   While data are available for most individual years, due to resource constraints we extracted data for only the "decade" years, except:

    -   also 1917, 1918, and 1919 to visualize the impact of the influenza pandemic at that time
    
    -   also 1929, because 1930 was not available.

    -   also 1939, because nuance of the available 1940 data prevented mapping to our standardized condition lists

    -   also 2018 and 2019 to visualize the impact of COVID-19

-   Documentation of all data sources is included as Appendix Table 1 below and is available here [hyperlink to be added eventually]

-   The systems for classifying deaths have evolved greatly over the years, from ad hoc systems, to "ICD-1" in the 1910's, ICD-8 in the 1960's, etc., to ICD-10 currently.

-   Therefore, we reviewed all the listed causes for each year, identified causes that occurred frequently in any one year, and developed a "common denominator" list of causes that, as much as possible, provided a single detailed list of specific cases to which all main causes of death from all years could be mapped.We then mapped/grouped this "Specific Level" grouping into a "Display Level" grouping with just 17 categories for increased comprehensibility and for visual display.

-   This "Display Level" was then grouped further into a "Top Level" for the broadest view of causes of death, and for consistency with other national and international systems.

-   Population data for decennial years (i.e., years ending in "0") was extracted directly from the U.S. Census Bureau, whereas population data for non-decennial years was found through a variety of sources, including the NCHS and the U.S. Census Bureau. Crude death rates were calculated with the following formula: Crude death rate per 100,000 = (total # of deaths/total population)*100,000.

-   All data processing, analysis, visualization, and production of this document were conducted using the R language and environment for statistical computing 

<br>

# Selected Findings:

-   Crude death rates have decreased substantially over the past 120 years, from 1,516 deaths per 100,000 population in 1900 to 801 deaths per 100,000 persons in 2010.  [Furthermore, since the age distribution of the population has shifted markedly to older ages over this period, the decrease in the age-adjusted mortality rate, a better overall measure of a population’s health, is proportionally much larger than the decrease in the crude rates--but age-adjusted rates cannot be calculated with the data from this project.] 
-   The crude death rate plots --- particularly the line plots --- should be interpreted carefully, as although the crude rate line plot shows that death rates for communicable disease have significantly decreased since the 1900's, a line plot of the number of deaths since 1900 reveals that though mortality from communicable disease trended downward during the early-mid 20th century, the HIV pandemic (around the 1980's) caused a surge in communicable disease mortality, as did the COVID-19 pandemic.
-  Since 1900,  causes of California mortality have shifted from communicable diseases to chronic diseases.  In 1990, the largest proportion of deaths, 8.08%, was due to communicable disease.  By 2019 this proportion had decreased to 2.66% of deaths, with 83.34% due to chronic conditions (including cancer and cardiovascular disease).
-   In 1900, the leading cause of death was Tuberculosis, followed by other communicable diseases (primarily communicable respiratory diseases). In contrast, Tuberculosis caused almost no deaths in 2019, and the leading causes, heart disease, Alzheimer’s disease and cancer are all chronic.

<br>

# Discussion:

-   This decrease in death rates can be somewhat attributed to advances in medical care, but perhaps even more important are the strides made through the work of public health professionals.
-   The emergence of COVID-19, a communicable disease, and one of the leading causes of death in 2020, clearly indicates that these patterns can and do change. 

<br>

<hr>

# SUMMARY VIEWS

## Figure 1. - Trend in death rates by Top Level conditions, by decade and additional pandemic years.

```{r fig.height= 8, fig.width= 15}

workGroup1a <- workDat0 %>% 
               group_by(year, topLevel, troubleshooting, displayLevel, specificLevel)     %>%  
               summarise(deaths=sum(Number)) %>%
  left_join(select(popDat, year, population), by = "year") %>%
  mutate(cDeathRate = 100000*deaths/population)

workGroup1b <- workGroup1a  %>%
  select(-population, -cDeathRate) %>%
               pivot_wider(names_from = year, values_from = deaths)  

# Crude Death Rates (topLevel)

###FOR MICHAEL
workGroupExplore  <- workDat0 %>%
                      group_by(year, topLevel, displayLevel, specificLevel)     %>%
                      summarise(deaths=sum(Number)) %>%
                      pivot_wider(names_from = year, values_from = deaths)


workGroup1a_plot_df <- workGroup1a %>%
  group_by(year, topLevel) %>%
  summarise(deaths = sum(deaths, na.rm = T),
            population = mean(population)) %>%
  ungroup() %>%
  mutate(cDeathRate = 100000*deaths/population) %>%
  mutate(barWidth = case_when(year %in% c(1939, 1940) ~ 0.9, 
                              year %in% c(1920, 1929, 2020) ~ 2, 
                              year %in% c(1900, 1910, 1950, 1960, 1970, 1980, 1990, 2000, 2010) ~ 6, 
                              year %in% c(1917, 1918, 1919, 2018) ~ 0.7))


ggplot(workGroup1a_plot_df, aes(x=year, y=cDeathRate, fill = topLevel, width = barWidth)) +
  geom_bar(colour = 'black', stat = 'identity') +
  scale_x_continuous(breaks = seq(1900, 2025, by = 10),limits = c(1895, 2025)) +
  labs(y = "Crude Death Rate per 100k Pop", x = "Cause of Death") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        axis.title.x = element_text(size=22),
        axis.title.y = element_text(size=22),
        legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'),
        legend.key.width = unit(1, 'cm'),
        legend.title = element_text(size=17),
        legend.text = element_text(size=13)
        )

```

<br>

<hr>

## Figure 2. - Trends in both number of deaths (2A) and crude death rates (2B) by Top Level conditions, 1900-2020 (excluding 1940), California

```{r fig.height= 7, fig.width= 12}

workGroup3 <- workDat0 %>% group_by(year, topLevel)  %>% 
               summarise(deaths=sum(Number)) %>%
  left_join(select(popDat, year, population), by = "year") %>%
  mutate(cDeathRate = 100000*deaths/population)


workGroup3_plot <- workGroup3 %>% filter(year != 1940)

# Number of Deaths Plot 
trendPlot_Ndeaths <- ggplot(workGroup3_plot, aes(x=year, y=deaths, color=topLevel)) +
  geom_line(size=2) +
  labs(title = "2A. Number of Deaths",
       y = "Deaths", x = "Year") +
  theme(plot.title = element_text(hjust=0.5),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))



# Crude Death Rate Plot 
trendPlot_cRates <- ggplot(workGroup3_plot, aes(x=year, y=cDeathRate, color=topLevel)) +
  geom_line(size=2) +
  labs(title = "2B. Crude Death Rate per 100,000", 
       y = "Deaths per 100,000", x = "Year") +
  theme(plot.title = element_text(hjust=0.5),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Top bottom
trendPlots_Ndeaths_cRates <- cowplot::plot_grid(
  trendPlot_Ndeaths + theme(legend.position = 'none'),
                                  NULL, 
  trendPlot_cRates + theme(legend.position = 'none'),
  ncol = 1, rel_heights = c(0.495, 0.01, 0.495))

# Get Legend
legend <- cowplot::get_legend(
  # create some space to the left of the legend
  trendPlot_cRates + theme(legend.box.margin = margin(0, 0, 0, 12), 
                           legend.position = "right", 
                           legend.title = element_text(size=17), 
                           legend.text = element_text(size=13))
)

# Final Plot
cowplot::plot_grid(trendPlots_Ndeaths_cRates, legend, nrow = 1, rel_widths = c(3, .5))


```

<br>

<hr>

## Figure 3. - Trend in death rates by Display Level conditions, by decade and additional pandemic years.

```{r fig.height= 8, fig.width= 15}

workGroup1b_plot_df <- workGroup1a %>%
  group_by(year, displayLevel) %>%
  summarise(deaths = sum(deaths, na.rm = T),
            population = mean(population)) %>%
  ungroup() %>%
  mutate(cDeathRate = 100000*deaths/population) %>%
  mutate(barWidth = case_when(year %in% c(1939, 1940) ~ 0.9, 
                              year %in% c(1920, 1929, 2020) ~ 2, 
                              year %in% c(1900, 1910, 1950, 1960, 1970, 1980, 1990, 2000, 2010) ~ 6, 
                              year %in% c(1917, 1918, 1919, 2018) ~ 0.7))

ggplot(workGroup1b_plot_df, aes(x=year, y=cDeathRate, fill = displayLevel, width = barWidth)) +
  geom_bar(colour = 'black', stat = 'identity') +
  scale_x_continuous(breaks = seq(1900, 2025, by = 10),limits = c(1895, 2025)) +
  labs(y = "Crude Death Rate per 100k Pop", x = "Cause of Death") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        axis.title.x = element_text(size=22),
        axis.title.y = element_text(size=22),
        legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'),
        legend.key.width = unit(1, 'cm'),
        legend.title = element_text(size=17),
        legend.text = element_text(size=13)
        )

```

<br>

<hr>

# DETAILED VIEWS

<hr>

## Figure 4. - Change in Display Level mortality between 1900 and 2020 grouped by Display Level.

```{r fig.height= 8, fig.width= 15}

# Number of Deaths Plot (specificLevel) - INCLUDE THIS CHART

ggplot(workGroup1a, aes(x=fct_rev(displayLevel), y=deaths, fill = topLevel)) +
  theme(axis.text.x = element_text(size=10, angle=90, hjust=0.5), axis.text.y = element_text(size=13), axis.title.x = element_text(size=22), axis.title.y = element_text(size=22)) +
  theme(legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'), 
        legend.key.width = unit(1, 'cm'), 
        legend.title = element_text(size=17), 
        legend.text = element_text(size=13)) +
  geom_col() +
  coord_flip() +
  labs(y = "Number of Deaths", x = "Cause of Death") +
  facet_wrap(vars(year),nrow=1,scales = "free_x")

```

<br>

<hr>

## Figure 5. - Crude death rates at the Display Level, by decade and additional pandemic years
  
```{r fig.height= 8, fig.width= 15}

# Crude Death Rates (displayLevel) - SHOULD I INCLUDE THIS CHART?

ggplot(workGroup1a, aes(x=fct_rev(displayLevel), y=cDeathRate, fill = topLevel)) +
  theme(axis.text.x = element_text(size=10, angle=90, hjust=0.5), axis.text.y = element_text(size=13), axis.title.x = element_text(size=22), axis.title.y = element_text(size=22)) +
  theme(legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'), 
        legend.key.width = unit(1, 'cm'), 
        legend.title = element_text(size=17), 
        legend.text = element_text(size=13)) +
  geom_col() +
  coord_flip() +
  labs(y = "Crude Death Rate per 100k Pop", x = "Cause of Death") +
  facet_wrap(vars(year),nrow=1)

```

<br>

<hr>

## Figure 6. - Change in Specific Level mortality between 1900 and 2020 grouped by Top Level.

```{r fig.height= 8, fig.width= 15}

# Number of Deaths Plot (specificLevel) - INCLUDE THIS CHART

ggplot(workGroup1a, aes(x=fct_rev(specificLevel), y=deaths, fill = topLevel)) +
  theme(axis.text.x = element_text(size=10, angle=90, hjust=0.5), axis.text.y = element_text(size=13), axis.title.x = element_text(size=22), axis.title.y = element_text(size=22)) +
  theme(legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'), 
        legend.key.width = unit(1, 'cm'), 
        legend.title = element_text(size=17), 
        legend.text = element_text(size=13)) +
  geom_col() +
  coord_flip() +
  labs(y = "Number of Deaths", x = "Cause of Death") +
  facet_wrap(vars(year),nrow=1,scales = "free_x")

```

<br>

<hr>

## Figure 7. - Crude death rates at the Specific Level, by decade and additional pandemic years

```{r fig.height= 8, fig.width= 15}

# Crude Death Rates (specificLevel)

ggplot(workGroup1a, aes(x=fct_rev(specificLevel), y=cDeathRate, fill = topLevel)) +
  theme(axis.text.x = element_text(size=10, angle=90, hjust=0.5), axis.text.y = element_text(size=13), axis.title.x = element_text(size=22), axis.title.y = element_text(size=22)) +
  theme(legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'), 
        legend.key.width = unit(1, 'cm'), 
        legend.title = element_text(size=17), 
        legend.text = element_text(size=13)) +
  geom_col() +
  coord_flip() +
  labs(y = "Crude Death Rate per 100k Pop", x = "Cause of Death") +
  facet_wrap(vars(year),nrow=1)

```

<br>

<hr>

# Key References

## Centers for Disease Control and Prevention. (2015, November 6). Products - vital statistics of the US - homepage. https://www.cdc.gov/nchs/products/vsus.htm. 

## Centers for Disease Control and Prevention. CDC WONDER. https://wonder.cdc.gov/. 

## United States Census Bureau. Explore Census Data. https://data.census.gov/cedsci/. 
```{r}
#library(gitcreds)
#gitcreds_set()


```
