---
title: "Century of Death"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message= FALSE, warning = FALSE)
```

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
library(forcats)
library(knitr)

y1900 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y1900") %>% mutate(year=1900)
y1910 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y1910") %>% mutate(year=1910)
y1917 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y1917") %>% mutate(year=1917)
y1918 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y1918") %>% mutate(year=1918)
y1920 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y1920") %>% mutate(year=1920)
y1929 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y1929") %>% mutate(year=1929)
y1939 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y1939") %>% mutate(year=1939)
y1940 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y1940") %>% mutate(year=1940)
y1950 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y1950") %>% mutate(year=1950)
y1960 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y1960") %>% mutate(year=1960)
y1970 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y1970") %>% mutate(year=1970)
y1990 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y1990") %>% mutate(year=1990)
y2000 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y2000") %>% mutate(year=2000)
y2010 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y2010") %>% mutate(year=2010)
y2018 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y2018") %>% mutate(year=2018)
y2019 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y2019") %>% mutate(year=2019)
y2020 <- read_excel("CA 1900-2020 Project.xlsx",sheet="y2020") %>% mutate(year=2020)



workDat0 <- bind_rows(y1900, y1910, y1917, y1918, y1920, y1929, y1939, y1940, y1950, y1960, y1970, y1990, y2000, y2010, y2018, y2019, y2020)



# Deaths and Population sheet
popDat <- read_excel("CA 1900-2020 Project.xlsx",sheet="Deaths and Popu.") %>%
  rename(year = 1, nDeaths = 2, population = 3, cDeathRate = 4) %>%
  select(-5)

#CCB Data Analysis
y2000ccb <- read_excel("ccb_cause_to_100_year_cause.xlsx",sheet="ccb_cause_to_100_year_causes",range="$C$1:$G$60") %>%
  rename(Number = deaths_2000) %>% 
  mutate(year=2000.2)
y2010ccb <- read_excel("ccb_cause_to_100_year_cause.xlsx",sheet="ccb_cause_to_100_year_causes")[,c(3,4,5,6,8)] %>%
  rename(Number = deaths_2010) %>% 
  mutate(year=2010.2)
y2018ccb <- read_excel("ccb_cause_to_100_year_cause.xlsx",sheet="ccb_cause_to_100_year_causes")[,c(3,4,5,6,9)] %>%
  rename(Number = deaths_2018) %>% 
  mutate(year=2018.2)



workDat1 <- bind_rows(y2000, y2000ccb, y2010, y2010ccb, y2018, y2018ccb)
```
# The 100+ years of California mortality project seeks to: 
## -Gather and standardize/sort data concerning California mortality from 1990 to present day
## -Analyze this data in order to identify patterns and create tools that will empower public health professionals to make more informed decisions.
## -Present the results of this project with useful visualizations (bar graphs, trend plots, etc.)

```{r fig.height= 8, fig.width= 15}
workGroup1a <- workDat0 %>% 
               group_by(year, topLevel, troubleshooting, specificLevel)     %>% 
               summarise(deaths=sum(Number)) %>%
  left_join(select(popDat, year, population), by = "year") %>%
  mutate(cDeathRate = 100000*deaths/population)

workGroup1b <- workGroup1a  %>%
  select(-population, -cDeathRate) %>%
               pivot_wider(names_from = year, values_from = deaths)  

# Number of Deaths Plot (specificLevel) - INCLUDE THIS CHART

ggplot(workGroup1a, aes(x=fct_rev(specificLevel), y=deaths, fill = topLevel)) +
  theme(axis.text.x = element_text(size=10, angle=90, hjust=0.5), axis.text.y = element_text(size=13), axis.title.x = element_text(size=22), axis.title.y = element_text(size=22)) +
  theme(legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'), 
        legend.key.width = unit(1, 'cm'), 
        legend.title = element_text(size=17), 
        legend.text = element_text(size=13)) +
  geom_col() +
  coord_flip() +
  labs(title = "Number of Deaths as Sorted by specificLevel",
       y = "Number of Deaths", x = "Cause of Death") +
  facet_wrap(vars(year),nrow=1,scales = "free_x") +
theme(plot.title = element_text(hjust = 0.5, size=30))

# Number of Deaths Plot (troubleshooting)

ggplot(workGroup1a, aes(x=fct_rev(troubleshooting), y=deaths, fill = topLevel)) +
  theme(axis.text.x = element_text(size=10, angle=90, hjust=0.5), axis.text.y = element_text(size=13), axis.title.x = element_text(size=22), axis.title.y = element_text(size=22)) +
  theme(legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'), 
        legend.key.width = unit(1, 'cm'), 
        legend.title = element_text(size=17), 
        legend.text = element_text(size=13)) +
  geom_col() +
  coord_flip() +
  labs(title = "Troubleshooting Deaths (Alternative to specificLevel)",
       y = "Number of Deaths", x = "Cause of Death") +
  facet_wrap(vars(year),nrow=1,scales = "free_x") +
theme(plot.title = element_text(hjust = 0.5, size=30))

# Crude Death Rates (specificLevel)

ggplot(workGroup1a, aes(x=fct_rev(specificLevel), y=cDeathRate, fill = topLevel)) +
  theme(axis.text.x = element_text(size=10, angle=90, hjust=0.5), axis.text.y = element_text(size=13), axis.title.x = element_text(size=22), axis.title.y = element_text(size=22)) +
  theme(legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'), 
        legend.key.width = unit(1, 'cm'), 
        legend.title = element_text(size=17), 
        legend.text = element_text(size=13)) +
  geom_col() +
  coord_flip() +
  labs(title = "Crude Death Rates (specificLevel)",
       y = "Crude Death Rate per 100k Pop", x = "Cause of Death") +
  facet_wrap(vars(year),nrow=1) +
  theme(plot.title = element_text(hjust = 0.5, size=30))

# Crude Death Rates (topLevel)

ggplot(workGroup1a, aes(x=year, y=cDeathRate, fill = topLevel)) +
  theme(axis.text.x = element_text(size=20, hjust=0.5), axis.text.y = element_text(size=20), axis.title.x = element_text(size=22), axis.title.y = element_text(size=22)) +
  theme(legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'), 
        legend.key.width = unit(1, 'cm'), 
        legend.title = element_text(size=17), 
        legend.text = element_text(size=13)) +
  geom_col(width = 0.9) +
  #coord_flip() +
  labs(title = "Crude Death Rates (topLevel)",
       y = "Crude Death Rate per 100k Pop", x = "Cause of Death") +
  theme(plot.title = element_text(hjust = 0.5, size=30))


```

<br>
<hr>


```{r fig.height= 8, fig.width= 12}
workGroup1c <- workDat1 %>% 
               group_by(year, topLevel, troubleshooting, specificLevel)     %>% 
               summarise(deaths=sum(Number)) %>%
  left_join(select(popDat, year, population), by = "year") %>%
  mutate(cDeathRate = 100000*deaths/population)

#CCB vs. CDC Specific Level (Graph)

ggplot(workGroup1c, aes(x=fct_rev(specificLevel), y=deaths, fill = topLevel)) +
  theme(axis.text.x = element_text(size=13, angle=90, hjust=0.5), axis.text.y = element_text(size=13), axis.title.x = element_text(size=22), axis.title.y = element_text(size=22)) +
  theme(legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'), 
        legend.key.width = unit(1, 'cm'), 
        legend.title = element_text(size=17), 
        legend.text = element_text(size=13)) +
  geom_col() +
  coord_flip() +
  labs(title = "Number of Deaths as Sorted by specificLevel",
       y = "Number of Deaths", x = "Cause of Death") +
  facet_wrap(vars(year),nrow=1) +
theme(plot.title = element_text(hjust = 0.5, size=25))

#CCB vs. CDC Specific Level (Table)

library(formattable)

customGreen0 = "#DeF7E9"
customGreen = "#71CA97"

workGroup1d <- workDat1  %>%
               group_by(year, specificLevel)     %>% 
               summarise(deaths=sum(Number)) %>%
  left_join(select(popDat, year, population), by = "year") %>%
  mutate(cDeathRate = 100000*deaths/population) %>%
  select(-population, -cDeathRate) %>%
               pivot_wider(names_from = year, values_from = deaths)

names(workGroup1d) <- c("specificLevel", "CDC2000", "CCB2000", "CDC2010", "CCB2010", "CDC2018", "CCB2018")

workGroup1d <- workGroup1d %>% mutate(DIF2000 = CDC2000-CCB2000)
workGroup1d <- workGroup1d %>% mutate(DIF2010 = CDC2010-CCB2010)
workGroup1d <- workGroup1d %>% mutate(DIF2018 = CDC2018-CCB2018)

formattable(workGroup1d,
            align =c("l","c","c","c","c", "c", "c"),
            list(`specificLevel` = formatter("span", style = ~ style(font.weight = "bold")),
                 'DIF2000'= color_tile(customGreen, customGreen0),
                 'DIF2010'= color_tile(customGreen, customGreen0),
                 'DIF2018'= color_tile(customGreen, customGreen0)
))

```
<br>
<hr>


```{r fig.height= 8, fig.width= 12}

workGroup3 <- workDat0 %>% group_by(year, topLevel)  %>% 
               summarise(deaths=sum(Number)) %>%
  left_join(select(popDat, year, population), by = "year") %>%
  mutate(cDeathRate = 100000*deaths/population)


# Number of Deaths Plot
ggplot(workGroup3, aes(x=year, y=deaths, color=topLevel)) +
  theme(axis.text.x = element_text(size=15), axis.text.y = element_text(size=15), axis.title.x = element_text(size=22), axis.title.y = element_text(size=22)) +
  geom_line(size=2) +
  theme(legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'), 
        legend.key.width = unit(1, 'cm'), 
        legend.title = element_text(size=17), 
        legend.text = element_text(size=13)) +
  labs(title = "Number of Deaths (topLevel Line Plot)",
       y = "Deaths", x = "Year") +
theme(plot.title = element_text(hjust = 0.5, size=25))

# Crude Death Rate Plot - INCLUDE THIS CHART

ggplot(workGroup3, aes(x=year, y=cDeathRate, color=topLevel)) +
 theme(axis.text.x = element_text(size=15), axis.text.y = element_text(size=15), axis.title.x = element_text(size=22), axis.title.y = element_text(size=22)) +
  geom_line(size=2) +
  theme(legend.key.size = unit(1, 'cm'), 
        legend.key.height = unit(1, 'cm'), 
        legend.key.width = unit(1, 'cm'), 
        legend.title = element_text(size=17), 
        legend.text = element_text(size=13)) +
  labs(title = "Crude Death Rate per 100,000 (topLevel Line Plot)",
       y = "Deaths per 100,000", x = "Year") +
theme(plot.title = element_text(hjust = 0.5, size=25))


```

<br>
<hr>

```{r}

#yNformat <- fmt_number(columns = 4:13, use_seps = TRUE, decimals = 0) 

workGroup1b %>% gt()  %>%
  tab_options(table.font.size = 9) %>%
  tab_style(
    style = list(cell_fill(color = "lightcyan"),
                 cell_text(weight = "bold")  ),
    locations =  cells_row_groups()    ) %>%
    fmt_number(columns = 4:12, use_seps = TRUE, decimals = 0) %>%
  summary_rows(columns= 4:12, fns = list(sum = ~sum(.,na.rm = TRUE)), formatter = fmt_number)
  


# https://stackoverflow.com/questions/66018397/how-to-specify-a-custom-formatter-function-gt-table-grand-summary



```





```{r}
#library(gitcreds)
#gitcreds_set()


```

